---
title: "Introduction to arbalistIO"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to arbalistIO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**arbalistIO** facilitates the import of single-cell ATAC-seq and Multiome (ATAC + Gene Expression) data into Bioconductor's `MultiAssayExperiment`. It provides efficient C++ based parsers for 10x Genomics fragment files to generate HDF5-backed sparse matrices for both genomic tiles and custom regions.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(arbalistIO)
library(MultiAssayExperiment)
library(SingleCellExperiment)
library(GenomicRanges)
library(rhdf5)
library(Matrix)
```

## 1. Preparing Mock Data

For this vignette, we will mock data that simulates the output of 10x Genomics Cell Ranger. This includes a fragment file (ATAC-seq) and a filtered feature matrix (RNA-seq).

### Generate a Fragment File

We use `mockFragmentFile` to create a fragment file containing reads for 50 cells.

```{r mock_fragment}
tmp_dir <- tempdir()
fragment_file <- file.path(tmp_dir, "fragments.tsv.gz")

seq_lengths <- c(chr1 = 10000, chr2 = 5000)
cell_names <- paste0("Cell", 1:50)

mockFragmentFile(
  output.file = fragment_file,
  seq.lengths = seq_lengths,
  num.fragments = 5000,
  cell.names = cell_names
)
```

### Generate a Mock Feature Matrix (HDF5)

`createArbalistMAE` requires the Cell Ranger filtered feature matrix (HDF5) to identify valid barcodes.


```{r mock_h5}
h5_file <- file.path(tmp_dir, "filtered_feature_bc_matrix.h5")

mockCellRangerH5(h5_file, 
                 n.cells=50,
                 cell.names = cell_names)

```

## 2. Creating SingleCellExperiment objects

We create a SingleCellExperiment object for the tile matrix of the ATAC-seq counts

```{r create_tile}
tile_sce <- createTileSCE(fragment.files=fragment_file, 
                          sample.names="MockSample",
                          tile.size=500, 
                          seq.lengths=seq_lengths)
                           
print(tile_sce)
```
We then create a SingleCellExperiment object for the region matrix of the ATAC-seq counts

```{r create_region}
promoters <- GRanges(
  seqnames = "chr1",
  ranges = IRanges(start = seq(100, 1000, by = 200), width = 100)
)

region_sce <- createRegionSCE(fragment.files=fragment_file, 
                              sample.names="MockSample",
                              regions=promoters)
                           
print(region_sce)
```

We next create a SingleCellExperiment object for the RNA-seq counts

```{r create_rna}
rna_sce <- createRNASCE(h5_file, sample.names="MockSample")
print(rna_sce)
```


## 3. Creating the MultiAssayExperiment

Now we can use `createArbalistMAE` to import everything into a single object. We first add the tile SCE and RNA SCE into MAE.

```{r create_mae}

mae <- createArbalistMAE(tile_sce, rna_sce)


print(mae)
```
We can append additional experiments to the existing MAE.

```{r add_sce}

mae <- c(mae, GeneAccessibilityMatrix = region_sce)
```
The resulting `MultiAssayExperiment` contains three experiments:

1.  **TileMatrix500**: 500bp tile counts (ATAC).
2.  **GeneAccessibilityMatrix**: Counts over the provided regions (ATAC).
3.  **GeneExpressionMatrix**: RNA counts imported from the HDF5 file.

### Exploring the Object

```{r explore_mae}
tiles <- experiments(mae)[["TileMatrix500"]]
print(tiles)

rna <- experiments(mae)[["GeneExpressionMatrix"]]
print(rna)

head(colData(mae))
```

## 4. Low-Level Matrix Generation

If you do not need the full `MultiAssayExperiment` structure, you can use the underlying C++ wrappers to generate specific matrices.

### Generating a Tile Matrix

Use `saveTileMatrix` to bin fragments into genome-wide tiles.

```{r save_tiles}
tile_h5 <- file.path(tmp_dir, "mytiles.h5")

tile_res <- saveTileMatrix(
  fragment.file = fragment_file,
  output.file = tile_h5,
  output.name = "counts",
  seq.lengths = seq_lengths,
  tile.size = 1000
)

# Returns a list with the GRanges tiles and the HDF5Matrix
tile_res$tiles
tile_res$counts
```

### Generating a Region Matrix

Use `saveRegionMatrix` to count fragments overlapping specific genomic regions

```{r save_regions}
region_h5 <- file.path(tmp_dir, "my_regions.h5")

# Define arbitrary regions
peaks <- GRanges("chr1", IRanges(start = c(10, 5000), width = 200))

region_mat <- saveRegionMatrix(
  fragment.file = fragment_file,
  output.file = region_h5,
  output.name = "counts",
  regions = peaks
)

print(region_mat)
```

That's it! For analyzing multiome datasets, check out the arbalist package.

## sessionInfo

```{r sessionInfo}
sessionInfo()
```
